---
title: "Dados sobre Arrecadação."
author: ""
date: ""
output: 
  html_document:
    highlight: espresso
    theme: united
---
#### `r format(Sys.Date(), '%d-%b-%Y')`. 
```{r setup, include=FALSE}
library(tidyverse)
library(readr)
library(lubridate)
library(forcats)
library(DT)
library(glue)
xfun::pkg_load2(c("htmltools", "mime"))

Sys.setlocale(category = "LC_ALL")
options(scipen = 999)

# load data in 'global' chunk so it can be shared by all users of the dashboard
  
arrec_ger_17_a_20_processada <- read_rds("arrec_ger_17_a_20_processada.rds")
serie_temp_arrec <- read_rds("serie_temp_arrec.rds")

graf_01 <- arrec_ger_17_a_20_processada %>% 
  #Retirando os meses de set, out, nov e dez.
  filter(mes < 9) %>% #Agrupando por ano para totalizar.
  group_by(ano) %>% 
  summarise(total_ate_agosto = sum(valor)/1e9) %>% 
  # Criando coluna com as diferenças percentuais,
  # tendo por base 2020.
  mutate(variacao_percentual = 
    round((total_ate_agosto[4]-total_ate_agosto)*100/total_ate_agosto[4], 1))

tot_2020 <- arrec_ger_17_a_20_processada %>% filter(ano == 2020) %>% 
group_by(tipo_de_arrecadacao) %>% summarise(val_por_trib = sum(valor, na.rm = TRUE)/1e9)

# Baixando a fonte Ubuntu.
library(showtext)
# Loading Google fonts (https://fonts.google.com/)
font_add_google("Ubuntu", "ubuntu")

# Automatically use showtext to render text
showtext_auto()

font <- "ubuntu"

```
## {.tabset .tabset-pills}
### __Arrecadação Geral Comparada__.

#### O gráfico mostra a arrecadação acumulada até agosto de 2020 (**`r glue('R$ {prettyNum(sum((arrec_ger_17_a_20_processada %>% filter(ano == 2020))$valor, na.rm = TRUE), big.mark = ".")}')`**) comparada com o mesmo período de 2017, 2018 e 2019. 
#### Os valores são em bilhões de R$. Está indicada a variação percentual em relação à 2020.

>#### Apesar da grave crise econômica causada pela pandemia, até agosto, a arrecadação de 2020 foi pouco abaixo (aprox. 5,3%) da de 2018 e 19.

###### __Obs.__ Toda análise foi feita com valores históricos.

```{r, echo=FALSE, fig.height=3, fig.width=9}

ggplot(graf_01, aes(x = ano, y = total_ate_agosto)) +
  geom_col(fill = "#dd4814") +
  geom_text(aes(label = prettyNum(round(total_ate_agosto, 1), decimal.mark = ",")), vjust = -0.3, hjust = 1.5, size=10, family="ubuntu") +
  geom_text(aes(label = paste0("(", prettyNum(round(variacao_percentual, 1), decimal.mark = ","), "%)")), vjust = -0.3,
                hjust = 0, size=10, family="ubuntu") +
  geom_hline(yintercept = 0, size = 1, colour="#333333") +
  ylim(0, 32) +
  labs(caption = "Fonte: CPAA / SUAR / SEFAZ") +
  ggplot2::theme(
    
    #Text format:
    #This sets the font, size, type and colour of text for the chart's title
    plot.title = ggplot2::element_text(family="ubuntu",
                                       size=30,
                                       face="bold",
                                       color="#222222"),
    #This sets the font, size, type and colour of text for the chart's subtitle, as well as setting a margin between the title and the subtitle
    plot.subtitle = ggplot2::element_text(family="ubuntu",
                                          size=22,
                                          margin=ggplot2::margin(9,0,9,0)),
    plot.caption = ggplot2::element_text(family="ubuntu",
                                         size=14),
    
    # Não gosto de legenda, então vou comentar (retirar) a do bbc_style e 
    # colocar legend.position = "none"
    #Legend format
    #This sets the position and alignment of the legend, removes a title and backround for it and sets the requirements for any text within the legend. The legend may often need some more manual tweaking when it comes to its exact position based on the plot coordinates.
    # legend.position = "top",
    # legend.text.align = 0,
    # legend.background = ggplot2::element_blank(),
    # legend.title = ggplot2::element_blank(),
    # legend.key = ggplot2::element_blank(),
    # legend.text = ggplot2::element_text(family=font,
    #                                     size=18,
    #                                     color="#222222"),
    legend.position = "none",
    
    
    #Axis format
    #This sets the text font, size and colour for the axis test, as well as setting the margins and removes lines and ticks. In some cases, axis lines and axis ticks are things we would want to have in the chart - the cookbook shows examples of how to do so.
    axis.title = ggplot2::element_blank(),
    axis.text = ggplot2::element_text(family="ubuntu",
                                      size=25,
                                      color="#222222"),
    axis.text.x = ggplot2::element_text(margin=ggplot2::margin(5, b = 5)),
    axis.ticks = ggplot2::element_blank(),
    axis.line = ggplot2::element_blank(),
    
    # Também acho que valores no eixo y poluem o gráfico.
    # Retirá-los-ei do gráfico.
    axis.title.y = ggplot2::element_blank(),
    axis.text.y = ggplot2::element_blank(),
    
    #Grid lines
    #This removes all minor gridlines and adds major y gridlines. In many cases you will want to change this to remove y gridlines and add x gridlines. The cookbook shows you examples for doing so
    panel.grid.minor = ggplot2::element_blank(),
    panel.grid.major.y = ggplot2::element_line(color="#cbcbcb"),
    panel.grid.major.x = ggplot2::element_blank(),
    
    #Blank background
    #This sets the panel background as blank, removing the standard grey ggplot background colour from the plot
    panel.background = ggplot2::element_blank(),
    
    #Strip background (#This sets the panel background for facet-wrapped plots to white, removing the standard grey ggplot background colour and sets the title size of the facet-wrap title to font size 22)
    strip.background = ggplot2::element_rect(fill="white"),
    strip.text = ggplot2::element_text(size  = 22,  hjust = 0)
  )

```

### __Variação da arrecadação: 2020/17, /18 e /19.__

#### Os gráficos indicam as variações das diferenças (em %) entre 2020 e 17; 2020 e 18; e, 2020 e 19. 
>#### A partir de ago/set, as linhas dos 3 gráficos são ascendentes, revelam uma reação positiva da arrecadação de 2020 frente aos outros anos (17, 18 e 19).

###### __Obs.__ Toda análise foi feita com valores históricos.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=9}

# Deixo abaixo os scripts que costumam ser mais trabalhosos.
# Lembrando a cor Ubuntu e os detalhes do geom_text.
# geom_col(fill = "#dd4814")

# Começando o gráfico.
serie_temp_arrec %>% ggplot(aes(mes, percentual)) + 
  geom_smooth(method = "loess", se = FALSE, size = 2, color = "#dd4814") +
  labs(caption = "Fonte: CPAA / SUAR / SEFAZ") +
  facet_wrap(~ var_percentual, ncol = 3) +
  
   # Aqui os textos estão na mesma altura e lado a lado, vjust iguais.
  # geom_text(aes(label = prettyNum(round(variavel_01, 1),
  # decimal.mark = ",")), vjust = -0.3, hjust = 1.5, size=10, family="ubuntu") +
  # geom_text(aes(label = paste0("(", prettyNum(round(variável_02, 1),
  # decimal.mark = ","), "%)")), vjust = -0.3, hjust = 0, size=10,  family="ubuntu") +
  
  # # Aqui os textos estão um sobre o outro, hjust iguais.
  # geom_text(aes(label = prettyNum(round(variavel_03, 3), decimal.mark = ",")),  vjust = -1.8, hjust = 0.5,   family="ubuntu", size=10) +
  # geom_text(aes(label = glue('{prettyNum(round(variavel_04*100/arrec_tot, 1), decimal.mark=",")}%')), vjust = -0.2, hjust = 0.5, family="ubuntu", size=10)
  
 # Para realçar o eixo dos x. 
  geom_hline(yintercept = 0, size = 1, colour="#333333") +
  scale_y_continuous(label = scales::label_percent()) +
  scale_x_continuous(breaks=c(1,3,5,7,9), 
                     labels=c("jan", "mar", "mai", "jul", "set")) +
  theme(
    
    #Text format:
    #This sets the font, size, type and colour of text for the chart's title
    plot.title = element_text(family="ubuntu",
                                       size=30,
                                       face="bold",
                                       color="#222222"),
    #This sets the font, size, type and colour of text for the chart's subtitle, as well as setting a margin between the title and the subtitle
    plot.subtitle = element_text(family="ubuntu",
                                          size=22,
                                          margin=margin(9,0,9,0)),
    plot.caption = element_text(family="ubuntu",
                                         size=14),
    
    # Não gosto de legenda, então vou comentar (retirar) a do bbc_style e 
    # colocar legend.position = "none"
    #Legend format
    #This sets the position and alignment of the legend, removes a title and backround for it and sets the requirements for any text within the legend. The legend may often need some more manual tweaking when it comes to its exact position based on the plot coordinates.
    # legend.position = "top",
    # legend.text.align = 0,
    # legend.background = element_blank(),
    # legend.title = element_blank(),
    # legend.key = element_blank(),
    # legend.text = element_text(family=font,
    #                                     size=18,
    #                                     color="#222222"),
    legend.position = "none",
    
    
    #Axis format
    #This sets the text font, size and colour for the axis test, as well as setting the margins and removes lines and ticks. In some cases, axis lines and axis ticks are things we would want to have in the chart - the cookbook shows examples of how to do so.
    axis.title = element_blank(),
    axis.text = element_text(family="ubuntu",
                                      size=25,
                                      color="#222222"),
    axis.text.x = element_text(margin=margin(5, b = 5)),
    # axis.ticks = element_blank(),
    # axis.line = element_blank(),
    
    # Também acho que valores no eixo y poluem o gráfico.
    # Retirá-los-ei do gráfico.
    # axis.title.y = element_blank(),
    # axis.text.y = element_blank(),
    
    #Grid lines
    #This removes all minor gridlines and adds major y gridlines. In many cases you will want to change this to remove y gridlines and add x gridlines. The cookbook shows you examples for doing so
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(color="#cbcbcb"),
    panel.grid.major.x = element_line(color="#cbcbcb"),
    
    #Blank background
    #This sets the panel background as blank, removing the standard grey ggplot background colour from the plot
    panel.background = element_blank(),
    
    #Strip background (#This sets the panel background for facet-wrapped plots to white, removing the standard grey ggplot background colour and sets the title size of the facet-wrap title to font size 22)
    strip.background = element_rect(fill="white"),
    strip.text = element_text(size  = 25,  hjust = 0)
  )

  
```


### __Arrecadação comparada por tributo__.

#### A comparação feita entre a arrecadação de 2020 e as anteriores (até agosto de cada ano), pode ser discriminada por tipo de arrecadação.

#### Em 'Search:' escreva o tipo de arrecadação que deseja comparar.

###### __Obs.__ Toda análise foi feita com valores históricos.

```{r, echo=FALSE, message=FALSE, fig.height=3, fig.width=9}

ty <- arrec_ger_17_a_20_processada %>% filter(mes < 9) %>% 
  group_by(ano, tipo_de_arrecadacao) %>%
  summarise(total = sum(valor, na.rm = TRUE)/1e9) %>%
  ungroup() %>% 
  mutate(ano = as.character(ano)) 

dp <- arrec_ger_17_a_20_processada %>% filter(mes < 9) %>% 
  filter(ano ==2020) %>% group_by(tipo_de_arrecadacao) %>% 
  summarise(var_temp = sum(valor, na.rm = TRUE)/1e9) %>%
  ungroup() 

inner_join(ty, dp) %>% mutate(delta_percentual = (var_temp - total)/var_temp) %>% select(c(1,2,3,5)) %>% 
   datatable(filter = 'none', caption = "Fonte: CPAA / SUAR / SEFAZ",
              rownames = FALSE, class = 'cell-border stripe',
              colnames = c('Ano', 'Tipo de arrecadação', 
                           'Valor (em bilhões de R$', 'Diferença %'),
              options = list(bPaginate = TRUE, pageLength = 5)) %>%
  formatCurrency(3,
  currency = "",
  interval = 3,
  mark = "",
  digits = 3,
  dec.mark = ",",
  before = TRUE) %>%
  formatPercentage(columns = 4, digits = 1, mark = ",") %>% 
  formatStyle(columns = 1:4, fontSize = '150%')
  
```

### __Arrecadação por tributo__.
#### O gráfico mostra a arrecadação até agosto de 2020 em valores absolutos (em bilhões de R$) e relativos.

#### A arrecadação com ICMS é 3,5 vezes maior que o total das restantes.

###### __Obs.__ Toda análise foi feita com valores históricos.

```{r, echo=FALSE, message=FALSE, fig.height=3, fig.width=9}

arrec_tot <- sum((arrec_ger_17_a_20_processada %>% filter(ano == 2020))$valor, na.rm = TRUE)/1e9

arrec_ger_17_a_20_processada %>% #Considerando apenas 2020.
  filter(ano == 2020) %>%
  group_by(tipo_de_arrecadacao) %>%
  summarise(tot_trib = sum(valor, na.rm = TRUE)/1e9) %>%
  ungroup() %>% # Ordenando os tributos por seus totais.
  mutate(tipo_de_arrecadacao = fct_reorder(tipo_de_arrecadacao, - tot_trib)) %>%
  mutate(arrec_percentual = tot_trib*100/arrec_tot) %>% 
  ggplot(aes(x = tipo_de_arrecadacao, y = tot_trib)) +
  geom_col(fill = "#dd4814") +
  geom_text(aes(label = prettyNum(round(tot_trib, 3), decimal.mark = ",")), vjust = -1.8, hjust = 0.5,   family="ubuntu", size=10) +
  geom_text(aes(label = glue('{prettyNum(round(tot_trib*100/arrec_tot, 1), decimal.mark=",")}%')), vjust = -0.2, hjust = 0.5, family="ubuntu", size=10) +
  geom_hline(yintercept = 0, size = 1, colour="#333333") +
  ylim(0, 28) +
  scale_x_discrete(labels=c("Dívida Ativa" = "Dív. Ativ.")) +
  labs(caption = "Fonte: CPAA / SUAR / SEFAZ") +
  
  theme(
    
    #Text format:
    #This sets the font, size, type and colour of text for the chart's title
    plot.title = ggplot2::element_text(family=font,
                                       size=28,
                                       face="bold",
                                       color="#222222"),
    #This sets the font, size, type and colour of text for the chart's subtitle, as well as setting a margin between the title and the subtitle
    plot.subtitle = ggplot2::element_text(family=font,
                                          size=22,
                                          margin=ggplot2::margin(9,0,9,0)),
    plot.caption = ggplot2::element_text(family=font,
                                         size=14),
    
    # Não gosto de legenda, então vou comentar (retirar) a do bbc_style e 
    # colocar legend.position = "none"
    #Legend format
    #This sets the position and alignment of the legend, removes a title and backround for it and sets the requirements for any text within the legend. The legend may often need some more manual tweaking when it comes to its exact position based on the plot coordinates.
    # legend.position = "top",
    # legend.text.align = 0,
    # legend.background = ggplot2::element_blank(),
    # legend.title = ggplot2::element_blank(),
    # legend.key = ggplot2::element_blank(),
    # legend.text = ggplot2::element_text(family=font,
    #                                     size=18,
    #                                     color="#222222"),
    legend.position = "none",
    
    
    #Axis format
    #This sets the text font, size and colour for the axis test, as well as setting the margins and removes lines and ticks. In some cases, axis lines and axis ticks are things we would want to have in the chart - the cookbook shows examples of how to do so.
    axis.title = ggplot2::element_blank(),
    axis.text = ggplot2::element_text(family=font,
                                      size=28,
                                      color="#222222"),
    axis.text.x = ggplot2::element_text(margin=ggplot2::margin(5, b = 10)),
    axis.ticks = ggplot2::element_blank(),
    axis.line = ggplot2::element_blank(),
    
    # Também acho que valores no eixo y poluem o gráfico.
    # Retirá-los-ei do gráfico.
    axis.title.y = ggplot2::element_blank(),
    axis.text.y = ggplot2::element_blank(),
    
    #Grid lines
    #This removes all minor gridlines and adds major y gridlines. In many cases you will want to change this to remove y gridlines and add x gridlines. The cookbook shows you examples for doing so
    panel.grid.minor = ggplot2::element_blank(),
    panel.grid.major.y = ggplot2::element_line(color="#cbcbcb"),
    panel.grid.major.x = ggplot2::element_blank(),
    
    #Blank background
    #This sets the panel background as blank, removing the standard grey ggplot background colour from the plot
    panel.background = ggplot2::element_blank(),
    
    #Strip background (#This sets the panel background for facet-wrapped plots to white, removing the standard grey ggplot background colour and sets the title size of the facet-wrap title to font size 22)
    strip.background = ggplot2::element_rect(fill="white"),
    strip.text = ggplot2::element_text(size  = 22,  hjust = 0)
  )
  
```

